Effectful Semantics for Substructural SSA

### 2023-11-14 pldi24/paper.tex
Tap

\begin{definition}[Distributive (pre)monoidal category] 
  A \textbf{distributive (pre)monoidal category} is a (pre)monoidal category with all finite coproducts equipped with a natural family of central isomorphisms (the \textbf{distributor}) \(\delta_{A, B, C}: (A + B) \otimes C \simeq A \otimes C + B \otimes C\). This isomorphism is subject to rather complex coherence conditions, worked out in full for the more general case of \textit{rig categories} in \cite{laplaza-distributivity}, but for our purposes we will only make use of the following:
  \begin{equation*}
    \delta_{A, B, C}^{-1} = [\iota_0 \otimes C, \iota_1 \otimes C] 
    \qquad
    \sigma^+_{A, B} \otimes C;\delta_{B, A, C} = \delta_{A, B, C};\sigma^+_{A \otimes C, B \otimes C}
  \end{equation*}
  Note in particular that this implies that \(\iota_i \otimes C;\delta_{A, B, C} = \iota_i\). A substructural (pre)monodial category is \textbf{distributive} if it is distributive as a (pre)monoidal category, and its distributor is \textit{pure}. A substructural effectful category \(\mc{C}\) is \textbf{distributive} if \(\mc{C}_0\) is distributive (as a substructural (pre)monoidal category).
\end{definition}
Many (pre)monoidal categories are distributive, including of course \(\ms{Set}\), but also categories which can be derived from monads on more primitive categories such as \(\ms{PFun}\) and \(\ms{Rel}\), as can be deduced from the following proposition:
\begin{proposition}
  If \(\mc{C}\) is a distributive substructural monoidal category and \(\mb{T}\) is a strong monad on \(\mc{C}\), then the Kleisli category \(\mc{C}_{\mb{T}}\) is a distributive substructural effectful category (with affine and relevant objects and morphisms inherited from \(\mc{C}\))
\end{proposition}
\begin{proof}
  Let \(\delta_{A, B, C}: \mc{C}((A + B) \otimes C, A \otimes C + B \otimes C)\) denote \(\mc{C}\)'s distributor. We will write composition in \(\mc{C}\) as "\(;\)" and composition in the Kleisli category \(\mb{T}\) as "\(\gg\)".
  We will show that
  \(\upg{\delta_{A, B, C}}{} = \delta_{A, B, C};\ms{pure}: \mc{C}_{\mb{T}}((A + B) \otimes C, A \otimes C + B \otimes C)\) satisfies the desired properties of a distributor. In particular, it is:
  \begin{itemize}
    \item Central: this follows immediately from being the image of a pure morphism
    \item Affine: \(\upg{\delta}{} \gg \upg{\ms{drop}}{} = \upg{(\delta;\ms{drop})}{} = \upg{\ms{drop}}{}\)
    \item Relevant: \(\upg{\delta}{} \gg \upg{\ms{split}}{} = \upg{(\delta;\ms{split})}{} = \upg{(\ms{split};\delta \otimes \delta)}{} = \upg{\ms{split}}{} \gg \upg{\delta}{} \otimes \upg{\delta}{}\)
    \item Left inverse: \(\upg{\delta}{} \gg \upg{[\iota_0 \otimes C, \iota_1 \otimes C]}{} = \upg{\delta;[\iota_0 \otimes C, \iota_1 \otimes C]}{} = \ms{id}\)
    \item Right inverse: \(\upg{[\iota_0 \otimes C, \iota_1 \otimes C]}{} \gg \upg{\delta}{} = \upg{[\iota_0 \otimes C, \iota_1 \otimes C];\delta}{} = \ms{id}\)
    \item Symmetric: \(\upg{\sigma^+}{} \otimes C \gg \upg{\delta}{} = \upg{\sigma^+ \otimes C;\delta}{} = \upg{\delta;\sigma^+}{} = \upg{\delta}{} \gg \upg{\sigma^+}{}\)
  \end{itemize}
\end{proof}

@@

\section{Appendix}

@@

\TODO{pull down to appendix? just trivial induction.}

>>>>

\begin{definition}[Distributive (pre)monoidal category] 
  A \textbf{distributive (pre)monoidal category} is a (pre)monoidal category with all finite coproducts equipped with a natural family of central isomorphisms (the \textbf{distributor}) \(\delta_{A, B, C}: (A + B) \otimes C \simeq A \otimes C + B \otimes C\). This isomorphism is subject to rather complex coherence conditions, worked out in full for the more general case of \textit{rig categories} in \cite{laplaza-distributivity}, but for our purposes we will only make use of the following:
  \begin{equation*}
    \delta_{A, B, C}^{-1} = [\iota_0 \otimes C, \iota_1 \otimes C] 
    \qquad
    \sigma^+_{A, B} \otimes C;\delta_{B, A, C} = \delta_{A, B, C};\sigma^+_{A \otimes C, B \otimes C}
  \end{equation*}
  Note in particular that this implies that \(\iota_i \otimes C;\delta_{A, B, C} = \iota_i\). A substructural (pre)monodial category is \textbf{distributive} if it is distributive as a (pre)monoidal category, and its distributor is \textit{pure}. A substructural effectful category \(\mc{C}\) is \textbf{distributive} if \(\mc{C}_0\) is distributive (as a substructural (pre)monoidal category).
\end{definition}
Many (pre)monoidal categories are distributive, including of course \(\ms{Set}\), but also categories which can be derived from monads on more primitive categories such as \(\ms{PFun}\) and \(\ms{Rel}\), as can be deduced from the following proposition:
\begin{proposition}
  If \(\mc{C}\) is a distributive substructural monoidal category and \(\mb{T}\) is a strong monad on \(\mc{C}\), then the Kleisli category \(\mc{C}_{\mb{T}}\) is a distributive substructural effectful category (with affine and relevant objects and morphisms inherited from \(\mc{C}\))
\end{proposition}
\begin{proof}
  Let \(\delta_{A, B, C}: \mc{C}((A + B) \otimes C, A \otimes C + B \otimes C)\) denote \(\mc{C}\)'s distributor. We will write composition in \(\mc{C}\) as "\(;\)" and composition in the Kleisli category \(\mb{T}\) as "\(\gg\)".
  We will show that
  \(\upg{\delta_{A, B, C}}{}: \mc{C}_{\mb{T}}((A + B) \otimes C, A \otimes C + B \otimes C)\) satisfies the desired properties of a distributor. In particular, it is:
  \begin{itemize}
    \item Central: this follows immediately from being the image of a pure morphism
    \item Affine: \(\upg{\delta}{} \gg \upg{\ms{drop}}{} = \upg{(\delta;\ms{drop})}{} = \upg{\ms{drop}}{}\)
    \item Relevant: \(\upg{\delta}{} \gg \upg{\ms{split}}{} = \upg{(\delta;\ms{split})}{} = \upg{(\ms{split};\delta \otimes \delta)}{} = \upg{\ms{split}}{} \gg \upg{\delta}{} \otimes \upg{\delta}{}\)
    \item Left inverse: \(\upg{\delta}{} \gg \upg{[\iota_0 \otimes C, \iota_1 \otimes C]}{} = \upg{\delta;[\iota_0 \otimes C, \iota_1 \otimes C]}{} = \ms{id}\)
    \item Right inverse: \(\upg{[\iota_0 \otimes C, \iota_1 \otimes C]}{} \gg \upg{\delta}{} = \upg{[\iota_0 \otimes C, \iota_1 \otimes C];\delta}{} = \ms{id}\)
    \item Symmetric: \(\upg{\sigma^+}{} \otimes C \gg \upg{\delta}{} = \upg{\sigma^+ \otimes C;\delta}{} = \upg{\delta;\sigma^+}{} = \upg{\delta}{} \gg \upg{\sigma^+}{}\)
  \end{itemize}
\end{proof}

@@

\section{Proofs}

\TODO{these}

\section{Rewriting}

@@
